Understanding Structure and Selectivity in Nanostructured Lyotropic Liquid Crystal Membranes using Molecular Simulations

### 2020-02-26 Ben_Manuscripts/thesis/intro.tex
Beefing up outline

  \section{Membranes for Selective Aqueous Separations}

  Pressure drive membrane processes have become an increasingly useful tool for
  performing aqueous separations.
  \begin{itemize}
    \item Untreated water can be a very complex solution with particles ranging
    in size from microns down to nanometers
    \item Sediment, bacteria, algae, proteins, small organic molecules, and ions
    are all typical components of aqueous streams.
%    \item Their original intention was for desalination applications but they 
%    have since expanded their reach to a more diverse set of chemical separation
%    problems. 
  \end{itemize}

  % BJC: This should probably go up top
  % TFC membranes review -- "Composite reverse osmosis and nanofiltration membranes"
  Membrane design is completely dependent on the target particle separation. 
  \begin{itemize}
    \item We have summarized the uses of different classes of membrane separation 
    technologies in Figure~\ref{fig:size_regimes}
    \item They can be classified based on the size of the particle they reject.
    \item Microfiltration membranes contain 0.1-1.0 $\mu$m pores. (separate bacteria, pigment removal)
    \item Ultrafiltration (sugars, proteins, colloidal materials) pore size about 0.05 $\mu$m
    \item Nanofiltration have pores on the order of 1 nm in size
    \item Reverse osmosis membranes are dense amorphous polymers with no explicit pores
    \item These technologies are often used in series
  \end{itemize}

@@

  \subsection{How Small Molecule Membrane Separations Work}

  %\wijmans_solution-diffusion_1995
  Membrane permeation is the result of a chemical potential gradient $\frac{d\mu_i}{dx}$.
  \begin{itemize}
	  \item The flux of component $i$ is proportional to this gradient:
	  \begin{equation}
	    J_i = -L_i \frac{d\mu_i}{dx}
	  \end{equation}
	  where $L_i$ is a coefficient of proportionality.
	  \item Chemical potential gradients can be induced by concentration, pressure, 
	  temperature and electromotive force.
  \end{itemize}

@@

  In porous membrane architectures, water flux is modeled as laminar flow through 
  cylindrical pores.
  \begin{itemize}
    \item The derivation follows Hagen-Poiseuille flow with inclusion of morphological
    characteristics:
    \begin{equation}
    A = \frac{\varepsilon r_p^2}{8\mu\delta_m}
    \end{equation}
    where $\varepsilon$ is the surface porosity, $r_p$ the pore radius, $\delta_m$ the
    membrane thickness, and $\mu$ the solution viscosity.
    \item Separation is considered in terms of rejection. Solutes with radii smaller
    than the pore will be rejected. A distribution of pore sizes prevents perfect
    rejection.
  \end{itemize}

  In amorphous membranes, water flux is modeled using the solution-diffusion 
  model.
  \begin{itemize}
    \item It is hypothesized that water and solute molecules partition into the
    membrane and diffuse across the polymer matrix due to a chemical potential
    gradient and then desorb into the permeate.
    \item The solubility and diffusivity can be described  in terms of the 
    diffusive water and solute permeabilities ($P_w$ and $P_s$) respectively
    \item \begin{equation}
    A = \frac{P_wV_w}{\delta_mR_gT}
    \end{equation}
    where $V_w$ is molar volume of water, $R_g$ is the gas constant and $T$ is
    absolute temperature.
    \item Solute flux is modeled as:
    \begin{equation}
    J_s = \frac{P_s}{\delta_m}\Delta c_m
    \end{equation}
  \end{itemize}

  These equations provide a somewhat limited perspective on the molecular details
  of transport.

  \subsection{Membrane Separation Applications}

@@

  Unfortunately, it is a difficult task to align the self-assembled hexagonal
  mesophases into continuous pores that traverse the width of the membrane.
  \begin{itemize}
    \item Although the membranes have shown high experimental selectivity, their 
    flux has been very low due to misalignment of the hexagonal mesophases.
    \item There has been 
  \end{itemize}

  \begin{itemize}
    \item Ideal geometry for transport
    \item Go through history
    \item Highlight issues with synthesis
    \item Talk about alignment approaches by Xunda Feng.
  \end{itemize}

@@

  \begin{itemize}
    \item Easier to synthesize
    \item History of monomers used
  \end{itemize}

@@

  \begin{itemize}
    \item Why MD?
    \item Why HII phase?
    \item What work has been done already
  \end{itemize}

  Questions to be answered: %BJC: one major question per chapter
  \begin{enumerate}
    %\item How do we build and equilibrate these systems? 
    \item What is the nanoscopic structure?  %MRS: good instinct to focus on this one. Build and equilibrate is a 'how' that allows you to get at this more important question.  
    \item What factors affect solute transport?  %MRS: make ``factors'' more specific and physical.
    \item Can we estimate macroscopic properties?
    \item How can we learn mechanisms with minimal human intervention?  %BJC: i.e. machine learning methods
  \end{enumerate}

>>>>

  \section{Membranes for Aqueous Separations}

  %BJC: limiting to pressure driven aqueous membrane separations (so I don't have to get into electrodialysis, gas separations, dialysis etc.)
  Pressure driven membrane processes have become an increasingly useful tool for
  performing aqueous separations.
  \begin{itemize}
    \item Untreated water can be a very complex solution with particles ranging
    in size from microns down to nanometers
    \item Sediment, bacteria, algae, proteins, small organic molecules, and ions
    are all common components of aqueous streams.
  \end{itemize}

  % TFC membranes review -- "Composite reverse osmosis and nanofiltration membranes"
  \noindent Membrane design is completely dependent on the target particle separation. 
  \begin{itemize}
    \item They can be classified based on the size of the particle they reject.
    \item Microfiltration membranes contain 100-10000 nm pores. They can separate large
    particles like bacteria and protozoa.~\cite{ma_functionalized_2014}
    \item Ultrafiltration membranes have pore sizes about 5-500 nm and are useful for the 
    separation of sugars, proteins, viruses and colloidal materials.~\cite{davey_ultrafiltration_2009}
    \item Nanofiltration have pores on the order of 1 nm in size and can be used for the separation
    of small organic molecules and charged species.~\cite{van_der_bruggen_review_2003}
    \item Reverse osmosis membranes are dense amorphous polymers with no explicit pores. Their
    dense polymer architecture reject all solutes except water making them useful for 
    separating hydrated ions from water.~\cite{warsinger_review_2018}
    \item We have summarized the uses of different classes of membrane 
    technologies in Figure~\ref{fig:size_regimes}.
    \item To deal with complex aqueous streams, and to prevent excessive membrane fouling, 
    these technologies are often used in series, removing larger particles first.~\cite{goosen_fouling_2005}
  \end{itemize}

@@

  \noindent Membrane permeation is the result of a chemical potential gradient, $\frac{d\mu_i}{dx}$.
  \begin{itemize}
	  \item The flux of component $i$ is proportional to this gradient:
	  \begin{equation}
	    J_i = -L_i \frac{d\mu_i}{dx}
	  \end{equation}
	  where $L_i$ is a coefficient of proportionality.~\cite{wijmans_solution-diffusion_1995}
	  \item Chemical potential gradients can be induced by concentration, pressure, 
	  temperature and electromotive forces.
	  \item We will limit our discussion to concentration and pressure gradients.
  \end{itemize}

@@

  In porous membrane architectures, water flux is modeled as laminar flow through 
  cylindrical pores.
  \begin{itemize}
    \item The derivation follows Hagen-Poiseuille flow with inclusion of morphological
    characteristics:
    \begin{equation}
      A = \frac{\varepsilon r_p^2}{8\mu\delta_m}
    \end{equation}
    where $\varepsilon$ is the surface porosity, $r_p$ the pore radius, $\delta_m$ the
    membrane thickness, and $\mu$ the solution viscosity.
    \item Separation is typically characterized in terms of solute rejection, $R$. 
    \begin{equation}
      R = 1 - \frac{c_p}{c_f}
    \end{equation}
    where $c_p$ and $c_f$ are the concentrations of the rejected species in
    the permeate and feed respectively.
    \item Solutes with radii smaller than the pore will be rejected. A distribution of 
    pore sizes prevents perfect rejection.
  \end{itemize}

  \noindent In amorphous membranes, water flux is modeled using the solution-diffusion 
  model.
  \begin{itemize}
    \item It is hypothesized that water and solute molecules partition into the
    membrane and diffuse across the polymer matrix due to a chemical potential
    gradient and then desorb into the permeate.
    \item The solubility and diffusivity can be described in terms of the 
    diffusive water and solute permeabilities ($P_w$ and $P_s$) respectively
    \item \begin{equation}
    A = \frac{P_wV_w}{\delta_mR_gT}
    \end{equation}
    where $V_w$ is molar volume of water, $R_g$ is the gas constant and $T$ is
    absolute temperature.
    \item Solute flux is modeled as:
    \begin{equation}
    J_s = \frac{P_s}{\delta_m}\Delta c_m = B\Delta c_m
    \end{equation}
    where B is a solute permeability coefficient and \Delta c_m is the concentration
    difference across the selective membrane layer.
    \item Under these assumptions, rejection can be reformulated as
    \begin{equation}
    R = \frac{A}{A + B}
    \end{equation}
    which approaches unity as the solute flux tends towards zero.
  \end{itemize}

  Although there has been considerable focus on creating membranes with high
  permeabilities, there is a case to be made that shifting the focus to membrane
  selectivity may offer a more effective route towards lowering the cost of 
  membrane separations.
  \begin{itemize}
    \item Increasing permeabilities beyond those already achieved may have 
    only a small effect on energy requirements and capital costs. 
    \item Cohen-Tanugi et al. calculated that tripling membrane permeability
    could reduce energy consumption up to 15\% in a seawater reverse osmosis plant.~\cite{cohen-tanugi_quantifying_2014}
    \item Perhaps the biggest detriment to energy consumption is the practical
	need for multiple membrane passes in order to achieve a desired purity.
    \item Additionally, many membranes are not suited for high purity separations
    of small neutral species which requires chemical and capital costs for
    post-treatment
    \item Addressing these problems with highly selective separations may lead to 
    more gains.
  \end{itemize}

%  These equations provide a somewhat limited perspective on the molecular details
%  of transport.
%  \begin{itemize}
%    \item The parameters are experimental observables which allow one to 
%    speculate on the atomic-level 
%  \end{itemize}

  \subsection{Applications of Selective Membranes}

@@

  Unfortunately, it is a difficult task to align the self-assembled hexagonal
  mesophases into continuous pores that traverse the width of the membrane.
  \begin{itemize}
    \item Although the membranes have shown high experimental selectivity, their 
    flux has been very low due to misalignment of the hexagonal mesophases.
  \end{itemize}

%  Recently, Feng et al. has come up with two methods which can be used to 
%  macroscopically align the hexagonal mesophases. 
  Considerable recent efforts have made progress towards the macroscopic
  alignment of the hexagonal mesophases.
  \begin{itemize}
    \item Feng et al. leverage the magnetic anisotropy of the columnar \
    mesophases in order to control their alignment with a magnetic field~\cite{feng_scalable_2014}
    \item Feng et al. also used an approach caused soft confinement which
    takes advantage of the hexagonal columns preference towards anchoring
    perpendicular to either a PDMS or glass substrate.~\cite{feng_thin_2016}
    \item Finally, Feng et al. designed a third technique which uses a structure
    directing molecule in order to template the assembly of a fatty acid 
    molecule into ordered columnar phases.~\cite{feng_highly_2017}
  \end{itemize}

@@

  The bicontinuous cubic, or Q\textsubscript{I} phase, is a class 
  of nanostructured phases characterized by 3D interconnected pores.
  \begin{itemize}
    \item Aside from its tortuous pore architecture, it shares the 
    uniform size and complex topological features of the H\textsubscript{II} phase
    which lends itself to highly selective separations.
    \item Although water and solutes must follow a longer path in order to pass
    through Q\textsubscript{I} membranes, they are more practical to synthesize
    because the mesophases do not require alignment. 
  \end{itemize}

  The space group of the Q\textsubscript{I} phase configuration formed by
  monomer 1 is uncertain and thought to be either Pn3m or Ia3d.
  \begin{itemize}
    \item 6 Q phase architectures have been identified in small molecule
    amphiphile systems.~\cite{mariani_cubic_1988}
    \item Of these, only 4 are consistent with diffraction data generated
    by the gemini surfactant used here.~\cite{pindzola_cross-linked_2003}
    \item Q$^{230}$ (Ia3d), Q$^{224}$ (Pn3m), Q$^{229}$ (Im3m) or Q$^{212}$ (P4\textsubscript{3}32) phase of type I configuration
    \item The presence of $1 / \sqrt{6}$ and $1 / \sqrt{8}$ peaks rules out the
    Q$^{227}$ (Fd3m) and Q$^{223}$ (Pm3n) configurations.
    \item The Pn3m and Ia3d architectures are the most common~\cite{mariani_cubic_1988,wiesenauer_nanoporous_2012}
    \item The most likely phase is the type I Q$^{230}$ (Ia3d) because it is the
    most common phase observed between lamellar and hexagonal phases and the
    monomeric alkyltrimethylammonium salts used to synthesize the gemini LCs
    exhibit clear Ia3d symmetry.
  \end{itemize}

  Due to its more facile synthesis, there has been significantly more development
  of Q\textsubscript{I} phase-forming monomers.
  \begin{itemize}
    \item The first generation of Q\textsubscript{I} membranes was made by 
    the self-assembly of a gemini phosphonium monomer in water.~\cite{pindzola_cross-linked_2003}
    \item Hatakeyama et al. improved the industrial viability of Q\textsubscript{I} 
    phase membranes by using a gemini ammonium monomer which is both easier and cheaper
    to synthesize.~\cite{hatakeyama_nanoporous_2010}
    \item Free standing films of both first and second generation Q\textsubscript{I} phase 
    membranes cannot withstand high pressure. Because solution casting was ineffective, Zhou
    et. al used a hot-pressing method to make mechanically strong membranes which involves 
    heating the initial monomer mixture to 70\degree C and pressing it with 12 tons of force
    into a thick microporous, hydrophilic polymer support.~\cite{zhou_new_2007}
    \item The most recent generation of Q\textsubscript{I} membranes uses an imidazolium-based
    gemini LLC monomer which is capable of being solution cast into defect-free thin films on
    porous supports. This improvement resulted in a ten-fold increase in flux while retaining
    selectivities similar to earlier generations.~\cite{carter_glycerol-based_2012}
  \end{itemize}

  %BJC: figure with different generation Q1 monomers

  State of the art Q\textsubscript{I} membranes offer selectivities that
  are competitive with existing commercial technologies.
  \begin{itemize}
    \item When separating organic solutes from NaCl, Q\textsubscript{I}-phase
    membrane filtration experiments have shown selectivity 2--3 times higher than
    commercial RO and 6--12 times higher than commercial NF membranes
    \cite{dischinger_application_2017}.
    \item Water permeability is higher than commercial RO membranes but
    less than commercial NF. There is work being done to reduce the
    thickness of the selective layer in order to increase permeability. 
  \end{itemize}

@@

  %BJC: following two paragraphs taken from first paper.
  Our current understanding of the molecular details of LLC polymer membranes'
  nanostructure is not sufficient to be able to precisely design them for
  specific separations. Dischinger et al.~attempted to use an empirical model
  that correlates the physiochemical properties of the counterion used in a
  Q\textsubscript{I}-phase LLC membrane to solute rejection\cite{dischinger_effect_2017}.
  Although their model showed some qualitative agreement with experiment, the
  quality of fit of their model was limited due to complex solute-membrane
  interactions that could not easily be modeled. Additionally, they observed
  an unexpected discrepancy in the relationship between uncharged solute
  rejection and water permeability, which will require a more in-depth knowledge of
  the difference between solute and solvent transport.

  Over the past 20 years, H\textsubscript{II}-phase LLC polymer membrane
  studies have been limited primarily to the Na-GA3C11 monomer with some
  characterization done after minor structural modifications. For example, Resel
  et al.~varied the length of the monomer tails and the counterion used and
  observed its effect on pore spacing \cite{resel_structural_2000}. In a later
  study of rejection performance, it was shown that membranes formed by
  cross-linked Na-GA3C11 in the H\textsubscript{II} phase cannot separate solutes
  less than 1.2 nm in diameter because the pores are too large
  \cite{zhou_supported_2005}. We do not yet understand how to controllably reduce
  the effective pore size or how to tune the chemical environment in the
  nanopores of this or related materials for small molecule separations. The only
  source of predictive modeling for LLC systems have been macroscopic models that
  likely do not adequately describe transport at these length scales
  \cite{hatakeyama_water_2011}. Modeling with molecular detail could provide
  sufficient information about the mechanisms and chemical features to better
  inform experimental design of similar nanostructured membranes.

  A molecular-level understanding of the relationship between monomer structure
  and solute transport can help provide guidelines to reduce the
  large chemical space available to design monomers for creation of
  separation-specific membranes.
  \begin{itemize}
    \item Atomistic MD simulations can provide the required level of detail 
    (Figure \ref{fig:detailed_pore}), assuming the force fields are sufficiently accurate. 
    \item With such an atomistic model, we can directly observe
  molecular-level solute transport and suggest governing mechanisms. 
    \item We can also observe how the choice of head group interacts with solutes
    of interest. 
    \item In addition, we can interchange counterions which may influence both the pore size
    and the strength of the Donnan potential.
  \end{itemize}

  \begin{figure}[!htb]
  \centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=0.6\textwidth]{cartoon_pore.pdf}
		\caption{}~\label{fig:undetailed_pore}
	\end{subfigure}
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=0.6\textwidth]{detailed_pore.pdf}
		\caption{}~\label{fig:detailed_pore}
	\end{subfigure} 
    \caption{(a) Previous understanding of the LLC pores are essentially speculations 
    based on limited chemical and experimental data. (b) We use detailed molecular 
    modeling in this paper in order to appropriately model the pore's complex architecture,
    which is crucial to understanding the mechanism of solute transport. In both 
    pictures, the head group region is colored blue and the tail region is colored cyan.}~\label{fig:detail}
  \end{figure}

  Although the Q\textsubscript{I} phase shows the most promise for
  practical applications, the focus of this work will be on the H\textsubscript{II}
  phase.
  \begin{itemize}
    \item The H\textsubscript{II} phase is an easier to geometry to model
    and analyze.
    \item We also have detailed structural data which is necessary for validating
    our model.
    \item All of the analysis techniques we use can be equally adapted to
    the more complicated Q\textsubscript{I} phase geometry.
  \end{itemize}

  There are few molecular simulation studies which study structure and transport
  in LLC systems.
  \begin{itemize}
    \item Mondal et al. used MD simulations to study the self assembly of gemini 
    surfactants and observed the formation of H\textsubscript{II} and Q\textsubscript{I}
    phases depending on water content.\cite{mondal_self-assembly_2013}
    \item Mantha and Yethiraj as well as Roy et al. studied the dynamics of confined 
    water in these systems and showed orders of magnitude differences in their motion
    dependent on system geometry.~\cite{mantha_dynamics_2016,roy_water_2016}
    \item Jackson et al. as well as Manthan et al. have combined experiment with simulation
    in order to show how the choice of monomer head group counterion regulate water 
    dynamics.~\cite{jackson_ion-specific_2018,mantha_counterion-regulated_2018}
    \item Sakamoto et al. and Nada et al. observed the dynamics of water molecules
    and ions and hypothesized transport mechanisms in a simplified model of an LLC
    nanopore.~\cite{sakamoto_development_2018,nada_transport_2020}
  \end{itemize}

  Because there is relatively sparse coverage of these types of simulation systems in 
  the literature, we have built a detailed molecular model from the ground up. There 
  are four primary research questions that we will address in this work.
  \begin{enumerate}
    \item What is the nanoscopic structure?
    	\begin{itemize}
    	  \item Before we can try to understand the molecular mechanisms of solute
    	  transport, we need to ensure that we model the chemical environment
    	  within the nanopores in a way that is consistent with experiment.
    	  \item We use experimental structural data in order to validate our model.
    	\end{itemize}
    \item Which solute-membrane interactions have the greatest influence on transport rates?
    	\begin{itemize}
    	  \item After gaining a detailed picture of the nanopore structure, we can 
    	  feel confident that solutes in this system will experience the same 
    	  interactions which are present in a real system.
    	  \item We create independent systems for each of 20 small polar solutes 
    	  and observe transport mechanisms whose dominance is dependent on
    	  solute chemical functionality.
    	  \item We characterize three different trapping mechanisms which lead
    	  to subdiffusive transport behavior.
    	\end{itemize}
    \item Can we estimate experimentally-relevant macroscopic properties?
    	\begin{itemize}
    	  \item Using our qualitative understanding of the dominant trapping mechanisms,
    	  we develop stochastic time series models which we can use to mimic solute dynamic
    	  behavior on time scales orders of magnitude longer than our simulations. 
    	  \item We attempt to reproduce both quantitative and qualitative solute trajectory
    	  behavior on MD simulation-length timescales.
    	  \item We then show how we can use our most promsiing model in order to connect 
    	  microscopic transport to macroscopic flux and selectivity.
    	\end{itemize}
    \item How can we learn mechanisms with minimal human intervention?  %BJC: i.e. machine learning methods
    	\begin{itemize}
    	  \item Although our stochastic models show great promise, their development 
    	  require some qualitative and quantitative understanding of dominant transport
    	  mechanisms.
    	  \item In the final chapter, we use the infinite hidden Markov model in order
    	  to automatically detect and parameterize a unknown number of hidden dynamical modes
    	  exhibited by solute time series. 
    	  \item This more flexible approach allows us to both infer mechanisms based 
    	  on differences in dynamical behavior and generate stochastic trajectory realizations
    	  which we can use to predict flux and selectivity. 
    	\end{itemize}
  \end{enumerate}
